---
title: "mctoolsr examples"
author: "Jon Leff"
date: "July 7, 2015"
output: 
  html_document: 
    keep_md: yes
---

This document serves as a brief introduction to using **mctoolsr**. This document will go through getting the pro-package working and a few examples using the most popular functions.

### Getting and using **mctoolsr**

**mctoolsr** is available on Github at: https://github.com/leffj/mctoolsr

To use:

1. [Install git](https://git-scm.com/book/en/v2/Getting-Started-Installing-Git) if you haven't already
2. Open a terminal window and clone **mctoolsr** in a directory where you keep software on your machine
```
git clone https://github.com/leffj/mctoolsr.git
```
3. When using **mctoolsr** in an R script, source the "routine_analysis_functions.R" file in the mctoolsr/R directory. For example, I include the following line at the top of all my R scripts using **mctoolsr**:

```{r}
source('~/Software/mctoolsr/R/routine_analysis_functions.R')
```


### Examples

Note that the following examples use an example dataset taken from a study examining the bacterial communities associated with fruits and vegetables ([Leff et al. 2013][ref1]). You can find this dataset in the mctoolsr/examples directory.    


#### Loading OTU tables and metadata

You can load a taxon (i.e. OTU) table in biom format using the following approach. Note that filepaths are specific to your system, so they will likely have to be altered depending on where you cloned **mctoolsr** into.

One of the nice things about loading your data this way is that all the sample IDs will be matched between your taxon table and metadata so that they will be in the same order and any sample IDs not present in one or the other will be dropped.

You can optionally filter out samples of a specific type during this step, but this can also be done separately as shown here.

```{r}
tax_table_fp = '~/Software/mctoolsr/examples/fruits_veggies_taxon_table_wTax.biom'
map_fp = '~/Software/mctoolsr/examples/fruits_veggies_metadata.txt'

input = load_taxon_table(tax_table_fp, map_fp)
```

The loaded data will consist of three parts:

1. The taxon table itself: "data_loaded"
2. The metadata: "map_loaded"
3. The taxonomic classifiers (if provided in the biom file): "taxonomy_loaded"

Any of these components can be quickly accessed using the '$' sign notation as shown in the next example.  


#### Returning numbers of sequences per sample

This can be achieved simply by calculating column sums on the taxon table:

```{r}
sort(colSums(input$data_loaded))
```  


#### Rarefying

As you can see from the previous example, we can rarefy (i.e. normalize for variable sequence depths) to 1000 sequences per sample without losing any samples. This can be done using the following command:

```{r}
input_rar = single_rarefy(input, 1000)
colSums(input_rar$data_loaded)
```  


#### Summarize taxonomic relative abundances at a higher taxonomic level

It is useful to get a feel for the taxonomic composition of your samples early on in the exploratory data analysis process. This can quickly be done by calculating taxonomic summaries at higher taxonomic levels - in this case at the phylum level. The values represent the sum of all the relative abundances for OTUs classified as belonging to the indicated phylum. In this example just the first few phyla and samples are shown.

```{r}
tax_sum_phyla = summarize_taxonomy(input_rar, level = 2, report_higher_tax = FALSE)
tax_sum_phyla[1:5, 1:8]
```  


#### Calculating a dissimilarity matrix

For dissimilarity-based analyses such as ordinations and PERMANOVA, it is necessary to calculate a dissimilarity matrix. There is currently support for Bray-Curtis dissimilarities based on square-root transformed data. This is a widely used dissimilarity metric for these analyses, but others will be added as requested.

```{r}
dm = calc_dm(input_rar$data_loaded)
```  


#### Plotting an ordination

There are two ways to plot ordinations in **mctoolsr**. The multistep way is shown here, but there is also a shortcut using the `plot_nmds()` function.

```{r}
ord = calc_ordination(dm, 'nmds')
plot_ordination(input_rar, ord, 'Sample_type', 'Farm_type', hulls = TRUE)
```  


#### Filtering samples

It is easy to filter samples from your dataset in **mctoolsr**. You can specify to remove samples meeting a specified condition in the metadata or keep those samples. In the example below, lettuce samples are removed, and the ordination is plotted again.

```{r}
input_rar_filt = filter_data(input_rar, 'Sample_type', filter_vals = 'Lettuce')
dm = calc_dm(input_rar_filt$data_loaded)
ord = calc_ordination(dm, 'nmds')
plot_ordination(input_rar_filt, ord, 'Sample_type', 'Farm_type', hulls = TRUE)
```  


#### Filtering taxa

There are multiple taxa filtering options in **mctoolsr**. This example shows how to explore the proteobacteria sequences across the samples. Taxa can also be filtered based on their relative abundance.

```{r}
input_proteobact = filter_taxa_from_data(input, taxa_to_keep = 'p__Proteobacteria')
sort(colSums(input_proteobact$data_loaded))
input_proteobact_rar = single_rarefy(input_proteobact, 219)
plot_nmds(calc_dm(input_proteobact_rar$data_loaded), map = input_proteobact_rar$map_loaded, 
          color_cat = 'Sample_type')
```  


#### Taxa based exploration

It is often useful to determine the taxa driving differences between the community compositions of different sample types. This example shows one way to do this to determine taxa driving differences between sample types.

```{r}
tax_sum_families = summarize_taxonomy(input_rar_filt, level = 5, report_higher_tax = FALSE)
taxa_summary_by_sample_type(tax_sum_families, input_rar_filt$map_loaded, 
                            factor = 'Sample_type', filter_level = 0.05, test_type = 'KW')
```

This analysis demonstrates that Pseudomonadaceae and Sphingobacteriaceae tend to have higher relative abundances on mushrooms than spinach and strawberries. The p values are based on Kruskal-Wallis tests and two different corrections are reported to deal with the multiple comparisons (Bonferroni and FDR). Rare taxa are filtered out using the `filter_level` peramter. The values indicated under the sample types are mean relative abundances.    


#### Calculating mean dissimilarities

Sometimes it is necessary to calculate mean dissimilarities. This is important in cases where sample types are pseudoreplecated. This is not the case here, but this example demonstrates this functionality.

```{r}
dm = calc_dm(input_rar_filt$data_loaded)
dm_aggregated = calc_mean_dissimilarities(dm, input_rar_filt$map_loaded, 
                                          'Sample_Farming', return_map = TRUE)
ord = calc_ordination(dm_aggregated$dm, ord_type = 'nmds')
plot_ordination(dm_aggregated, ord, color_cat = 'Sample_type')
```



[ref1]: http://journals.plos.org/plosone/article?id=10.1371/journal.pone.0059310



